Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-08-03 paper/ms.tex
moving paper over from sharelatex

\linespread{0.54} % close to 10/13 spacing in ``manuscript''
\setlength{\parindent}{0.54\baselineskip}
\hypersetup{colorlinks = false}
\makeatletter % you know you are living your life wrong when you need to do this
\long\def\frontmatter@title@above{
\vspace*{-\headsep}\vspace*{\headheight}
\noindent\footnotesize
{\noindent\footnotesize\textsc{\@journalinfo}}\par
{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6\\
With modifications by David W. Hogg and Daniel Foreman-Mackey
}\par\vspace*{-\baselineskip}\vspace*{0.625in}
}%
\makeatother

@@

\author{%
    Eric~Agol\altaffilmark{\uw} and
    Daniel~Foreman-Mackey\altaffilmark{\uw,\sagan}
}

@@

Our Gaussian Processes are fast.

@@

This is the literature \citep{Ambikasaran:2015}.

@@

>>>>

\linespread{0.54} % close to 10/13 spacing in ``manuscript''
\setlength{\parindent}{0.54\baselineskip}
\hypersetup{colorlinks = false}
\makeatletter % you know you are living your life wrong when you need to do this
\long\def\frontmatter@title@above{
\vspace*{-\headsep}\vspace*{\headheight}
\noindent\footnotesize
{\noindent\footnotesize\textsc{\@journalinfo}}\par
{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6 with modifications
}\par\vspace*{-\baselineskip}\vspace*{0.625in}
}%
\makeatother

@@

\author{%
    Daniel~Foreman-Mackey\altaffilmark{\uw,\sagan} and
    Eric~Agol\altaffilmark{\uw}
}

@@

We derive a fast, $O(N)$, Gaussian Process approach that involves Lorentzian
kernels.  This basis set can approximate commonly used kernels, and is a good
description of stellar variability.

@@

\section{Outline}

\begin{enumerate}
\item Non-parametric modeling of time series with correlated noise using Gaussian processes
\item Assumptions:\\
  a). Stationary noise (although this can be broken).\\
  b). Gaussian
\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
   curves; Solar curves; short-cadence.
\item Solution(s):\\
  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
\item Problem: HODLR is tricky to use; CARMA requires certain relations between
  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
   $O(N*p^2)$, where p is number of Lorentzians.
\item Generalized Ambikasarn for complex coefficients (Hermitian).
\item Performance:\\
  a). Show $O(N*p^2)$ scaling\\
  b). Show comparison to HODLR.\\
  c). Show comparison to wavelet (ala Carter \& Winn).\\
  d). Compare with CARMA Kallman filter(?).
\item Various methods/formulae:\\
  a). Likelihood computation.\\
  b). Generating GP with particular covariance [yet to solve]. [ ]\\
  c). Inferring particular components.\\
  d). Computing derivatives of GPs (as used in RV methods).\\
  e). Forecasting/interpolating.\\
  f). Inferring confidence intervals.\\
  g). Derivative of likelihood function with respect to kernel parameters.\\
  h). Non-stationary GP (can the coefficients vary with time?).\\
  i). Multi-band (each element of extended matrix becomes a matrix).

\item Generalization to non-stationary noise.
\item Example application: TYC 3559\\
  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
  b). Analysis of solar data.

\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.

\item Appendix:
  a). Description/summary of mathematics.\\
  b). Description of code.
\end{enumerate}

\subsection{Derivative of likelihood with respect to Kernel parameters}

The derivative of the likelihood can be computed in $O(N)$ operations as follows.

The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
\begin{equation}
    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
\end{equation}
where $^\prime$ means the derivative wrt $x_i$.
The kernel may be written as:
\begin{equation}
    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
    \begin{array}{cc}
       w_i + \sum_{l=1}^p \alpha_l & j = i \\
       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
    \end{array}
    \right.
\end{equation}
Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.

Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
\begin{eqnarray}
    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
     \left\{
    \begin{array}{cc}
       1 & j = i \\
        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
    \end{array}
    \right.\\
    \frac{\partial K_{ij}}{\partial \beta_k} &=&
     \left\{
    \begin{array}{cc}
       0  & j = i \\
        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
    \end{array}
    \right.
\end{eqnarray}
The first term in the derivative of the likelihood can be computed in $O(N)$
evaluations, while the second term can be computed in three steps:
\begin{itemize}
\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
\item Next, create an e
\end{itemize}
Writing this term out:

@@

EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
NASA Astrobiology Institute’s Virtual Planetary Laboratory, supported
by NASA under cooperative agreement NNH05ZDA001C.
